Evolution of Processes in Delivering Reliable Software in the European Scientific Arena

### 2018-06-25 paper-ieee-trans.tex
all reviwer 3 comments done, remove some latex comments

{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
leveraged the lessons learned from the previous experiences described in
Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
profitable regarding the evaluation and application of innovative software engineering
methodologies, management of new technologies to put those methodologies
into practice, compilation and analysis of an appropriate set of metrics to measure the
quality of the implemented solutions.
An additional level of complexity, comes from the fact that these large collaborative projects
have an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
% Comment
%%P3 C2 L29: This introductory paragraph goes into a bit too much jargony detail, I think, especially given that it is followed by in-depth coverage of each "pillar". I suggest a clearer identification of each pillar, maybe via bullets or definition paragraphs, avoiding or with general-audience definition of jargon like DevOps, delivery mechanisms, two-factor validation, pilot testbeds, and early adoption. I appreciate this may all be highly meaningful to someone steeped in this operational area, but to an interested outsider it reads much like management jargon -- particuarly the final sentence: is this a long-winded way of saying "used during the commissioning period"?
software lifecycle:
\begin{enumerate}
    \item The definition of a \textit{Software Quality Assurance (SQA) criteria and a metrics gathering procedure} to provide guidelines
    aiming at developing quality software, validating each change in the code, to facilitate its adoption. The software lifecycle is
    continuously monitored by a set of metrics, allowing prompt reactions to detected malfunctions that may occur throughout the process. 
    \item The \textit{promotion of automation} to enable the foregoing change--based validation approach and to accelerate the delivery
    process of new software versions.
    \item A post-release validation to be carried out over the project's \textit{preview testbeds}, where the new product versions
    are deployed as part of the integration validation, and the \textit{testing in production environments} by external resource providers.
\end{enumerate}

@@

\subsection{Software quality validation and delivery automation}
\label{sec:devops}
% Comment
%P5 C2 L21: Define DevOps!
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more 
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}.

@@

Jenkins is an event--response CI and Docker container
provisioning system that provides instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
% Comment from reviewer
%Please add a reference.
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, that have been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs)
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre--production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a
testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On
successful completion, the packages are then automatically signed and subsequently moved to
the preview branch, from where they finally be made available in the production branch after
the release manager supervises the process.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

>>>>

{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
leveraged the lessons learned from the previous experiences described in
Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
profitable regarding evaluation and application of innovative software engineering
methodologies, management of new technologies to put those methodologies
into practice, compilation and analysis of an appropriate set of metrics to measure the
quality of the implemented solutions.
An additional level of complexity, comes from the fact that these large collaborative projects
have an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the software lifecycle:

\begin{enumerate}
    \item The definition of a \textit{Software Quality Assurance (SQA) criteria and a metrics gathering procedure} to provide guidelines
    aiming at developing quality software, validating each change in the code, to facilitate its adoption. The software lifecycle is
    continuously monitored by a set of metrics, allowing prompt reactions to detected malfunctions that may occur throughout the process. 
    \item The \textit{promotion of automation} to enable the foregoing change--based validation approach and to accelerate the delivery
    process of new software versions.
    \item A post-release validation to be carried out over the project's \textit{preview testbeds}, where the new product versions
    are deployed as part of the integration validation, and the \textit{testing in production environments} by external resource providers.
\end{enumerate}

@@

\subsection{Software quality validation and delivery automation}
\label{sec:devops}
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more 
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}.

@@

Jenkins is an event--response CI and Docker container
provisioning system that provides instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update.

The CI pipeline is composed from a set of required checks, their exit status
is reported back to GitHub. The changes can be prevented if those tests are
not successfully executed.
As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, that have been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs)
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre--production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a
testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On
successful completion, the packages are automatically signed and subsequently moved to
the preview branch.
In the last step, the release manager supervises the process to move the  packages to the
production branch.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.
